Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-22 survey/survey.tex
Introduce reversing combinator, as a derived combinator, in section 3 (as it is used in section 4). Then continue describing section 4 combinators.

Embedded in our definition of $\Pi$ is a conscious design decision: to make the
terms of $\Pi$ \emph{syntactically} reversible. In other words, to
every $\Pi$ constant, there is another $\Pi$ constant which is its
inverse. This is not the only choice.  Another would be to add a
$\mathit{flip}$ combinator to the language; we could then remove
quite a few combinators as redundant. The drawback is that many
programs in $\Pi$ become longer. Furthermore, some of the symmetry
at ``higher levels'' (see next section) is also lost. Since the
extra burden of language definition and of proofs is quite low, we
prefer the structural symmetry over a minimalistic language definition.

@@

\begin{figure}[tb]
Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\idc \odot \, c \Leftrightarrow c}
\quad
  {c \, \odot \idc \, \Leftrightarrow c}
\quad
  {c\,\, \odot\,! c \Leftrightarrow \idc}
\quad
  {! c \odot c \Leftrightarrow \idc}
\\
  {c \Leftrightarrow c}
\quad
\Rule{}
  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
  {c_1 \Leftrightarrow c_3}
  {}
\quad
\Rule{}
  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
  {}
\end{array}\]
\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
\end{figure}

The constructs in Fig.~\ref{figh} express various properties of composition.
The first two says that $\idc$ is a left and right identity for sequential composition.

>>>>

\begin{figure}[t]
\[
\Rule{}
{\jdg{}{}{c_1 : t_1 \iso t_2}}
{\jdg{}{}{\ !\ c_1 : t_2 \iso t_1}}
{}
\]
\caption{Derived $\Pi$-combinator.}
\label{derived-pi-combinators}
\end{figure}

Embedded in our definition of $\Pi$ is a conscious design decision: to make the
terms of $\Pi$ \emph{syntactically} reversible. In other words, to
every $\Pi$ constant, there is another $\Pi$ constant which is its
inverse. As this is used frequently, we give it the short name $!$,
and its type is given in Fig.~\ref{derived-pi-combinator}. This
combinator is \emph{defined}, by pattern matching on the syntax of
its argument and structural recursion.

This is not the only choice.  Another would be to add a
$\mathit{flip}$ combinator to the language; we could then remove
quite a few combinators as redundant. The drawback is that many
programs in $\Pi$ become longer. Furthermore, some of the symmetry
at ``higher levels'' (see next section) is also lost. Since the
extra burden of language definition and of proofs is quite low, we
prefer the structural symmetry over a minimalistic language definition.

@@

\begin{figure}[tb]
Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\idc \odot \, c \Leftrightarrow c}
\quad
  {c \, \odot \idc \, \Leftrightarrow c}
\quad
  {c\,\, \odot\,!\, c \Leftrightarrow \idc}
\quad
  {!\, c \odot c \Leftrightarrow \idc}
\\
  {c \Leftrightarrow c}
\quad
\Rule{}
  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
  {c_1 \Leftrightarrow c_3}
  {}
\quad
\Rule{}
  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
  {}
\end{array}\]
\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
\end{figure}

The constructs in Fig.~\ref{figh} express various properties of composition.
The first two says that $\idc$ is a left and right identity for sequential composition.
The next two say that all programs are reversible, both on the left and the right:
running $c$ and then its reverse ($!\, c$) is equivalent to the identity, and the
same for doing $!\, c$ first then $c$. The last line say that there is an
identity level-2 combinator, a sequential composition, and that level-2
equivalence respects level-1 sequential composition $\odot$.
